---
title: "The effect of education on income"
author: Marc Parcerisa
date: November 17, 2024
output: 
    md_document:
        variant: markdown_github
---

<!-- README.md is generated from IncomeStudy.Rmd. Please edit that document. -->

```{r opts, echo = FALSE}
knitr::opts_chunk$set(fig.path = "images/")
```

Today I saw some kid on *Tik Tok* saying that there's no point in studying anything
because doing so wouldn't increase your income. I thought that was a pretty bold
statement, so I did what any reasonable person would do: I went to Kaggle to find
a dataset that would help me prove him wrong.

I found a dataset called `income.csv` that contains information about american
people's income and education.

You can find it here: 
https://www.kaggle.com/datasets/amirhosseinmirzaie/americancitizenincome?resource=download

Here's an explanation of the columns:

**Column**     | **Description**
:-------------:|:-------------:
age            | Age
workclass      | A general term indicating the employment status of an individual.
fnlwgt         | Final weight, representing the number of individuals that this row represents (a representative sample).
education      | Highest level of education achieved by an individual.
education.num  | Highest level of education achieved by an individual in numerical form.
marital.status | Marital status of an individual. Note that Married-civ-spouse refers to a civilian spouse, and Married-AF-spouse refers to a spouse in the Armed Forces.
occupation     | General type of occupation of an individual.
relationship   | Relationship of this individual with others, for example, spouse (Husband). Each data point has only one relationship.
race           | Race
sex            | Biological sex of an individual.
capital.gain   | Capital gains of an individual.
capital.loss   | Capital losses of an individual.
hours.per.week | Number of hours the individual reported working per week.
native.country | Country of origin.
income         | Income, less than or equal to $50,000 (<=50K) or more than that (>50K).

For this study I'll be using the following libraries:

```{r}
library(readr)
library(FactoMineR)
```

### Loading and preparing the data

```{r}
df <- read_csv("income.csv")
summary(df)
```
